---
output: word_document
---
# Scripts to analyze Goby vertebral morphology data
Created by Kelly Diamond

### Preprocessing steps and notes
- fish standard length (length not including caudal fin) was measured in ImageJ
- centrum length (cl) and intervertebral space (ivs) was mesured for the vertebrae at 25/50/75% of the vertebral length. Lengths are calculatedin the script from landmark points placed using 3D slicer
- Measurments are size standardized by the standard length
- all measurments recorded in mm

```{r load libraries, include=FALSE}
library(data.table)
library(ggplot2)
library(grid) # necessary for ggplot2
library(gridExtra) #grid.arrange
library(doBy)
library(pracma) #for dot products
library(reshape2)
```

### Import data
The vert data table is for calculating distance measures and GobyData is the reorganized distances used in the analysis for this study. 
```{r import data, include=FALSE}
vert = data.table(read.table('slicer_output_22may19.txt',header=TRUE,sep='\t'))
GobyData = read.csv('./slicer_distance_measures_1april21.csv', header = TRUE)
GobyData$position = as.character(GobyData$position)
GobyData$style = as.character(GobyData$style)
```

#Distances
Calculate linear distances between points landmarked using 3D Slicer.
```{r calc dist, include=FALSE}
#function for calculating euclidian distances between 2 xyz pts
euc.dist <- function(x1, x2, y1, y2, z1, z2 ) as.numeric(sqrt( (x2 - x1)^2 + (y2 - y1)^2 + (z2 - z1)^2 )) #calculates distance between 2 xyz points
std.by.length <- function(x1,y1) as.numeric(x1/y1) #standardize a measure (x1) by standard length (y1)
dist.output <- as.data.table(vert[,1:4])
dist.output$std_length <- as.numeric(dist.output$std_length)

#for loop iterates though each individual and does all the math to calculate measures involving distance
for (i in 1:dim(vert) [1]) {

  #centrum length in mm
  dist.output$cl_25[i] <- euc.dist(vert$X25p_1x[i],vert$X25p_2x[i],vert$X25p_1y[i],vert$X25p_2y[i],vert$X25p_1z[i],vert$X25p_2z[i])
  dist.output$cl_50[i] <- euc.dist(vert$X50p_1x[i],vert$X50p_2x[i],vert$X50p_1y[i],vert$X50p_2y[i],vert$X50p_1z[i],vert$X50p_2z[i])
  dist.output$cl_75[i] <- euc.dist(vert$X75p_1x[i],vert$X75p_2x[i],vert$X75p_1y[i],vert$X75p_2y[i],vert$X75p_1z[i],vert$X75p_2z[i])
  
  #centrum length in body length (bl)
  dist.output$cl_25_std[i] <- std.by.length(dist.output$cl_25[i],dist.output$std_length[i])
  dist.output$cl_50_std[i] <- std.by.length(dist.output$cl_50[i],dist.output$std_length[i])
  dist.output$cl_75_std[i] <- std.by.length(dist.output$cl_75[i],dist.output$std_length[i])

  #minimum centrum width in mm
  dist.output$can.d_25[i] <- euc.dist(vert$X25p_4x[i],vert$X25p_5x[i],vert$X25p_4y[i],vert$X25p_5y[i],vert$X25p_4z[i],vert$X25p_5z[i])
  dist.output$can.d_50[i] <- euc.dist(vert$X50p_4x[i],vert$X50p_5x[i],vert$X50p_4y[i],vert$X50p_5y[i],vert$X50p_4z[i],vert$X50p_5z[i])
  dist.output$can.d_75[i] <- euc.dist(vert$X75p_4x[i],vert$X75p_5x[i],vert$X75p_4y[i],vert$X75p_5y[i],vert$X75p_4z[i],vert$X75p_5z[i])
  
  #minimum centrum width in bl
  dist.output$can.d_25_std[i] <- std.by.length(dist.output$can.d_25[i],dist.output$std_length[i])
  dist.output$can.d_50_std[i] <- std.by.length(dist.output$can.d_50[i],dist.output$std_length[i])
  dist.output$can.d_75_std[i] <- std.by.length(dist.output$can.d_75[i],dist.output$std_length[i])

  #maximum centrum width in mm
  dist.output$con.d_25[i] <- euc.dist(vert$X25p_2x[i],vert$X25p_3x[i],vert$X25p_2y[i],vert$X25p_3y[i],vert$X25p_2z[i],vert$X25p_3z[i])
  dist.output$con.d_50[i] <- euc.dist(vert$X50p_2x[i],vert$X50p_3x[i],vert$X50p_2y[i],vert$X50p_3y[i],vert$X50p_2z[i],vert$X50p_3z[i])
  dist.output$con.d_75[i] <- euc.dist(vert$X75p_2x[i],vert$X75p_3x[i],vert$X75p_2y[i],vert$X75p_3y[i],vert$X75p_2z[i],vert$X75p_3z[i])
  
  #maximum centrum width in bl
  dist.output$con.d_25_std[i] <- std.by.length(dist.output$con.d_25[i],dist.output$std_length[i])
  dist.output$con.d_50_std[i] <- std.by.length(dist.output$con.d_50[i],dist.output$std_length[i])
  dist.output$con.d_75_std[i] <- std.by.length(dist.output$con.d_75[i],dist.output$std_length[i])
  
  #intervertebral space in mm
  dist.output$ivs_25[i] <- euc.dist(vert$X25p_2x[i],vert$X25p_6x[i],vert$X25p_2y[i],vert$X25p_6y[i],vert$X25p_2z[i],vert$X25p_6z[i])
  dist.output$ivs_50[i] <- euc.dist(vert$X50p_2x[i],vert$X50p_6x[i],vert$X50p_2y[i],vert$X50p_6y[i],vert$X50p_2z[i],vert$X50p_6z[i])
  dist.output$ivs_75[i] <- euc.dist(vert$X75p_2x[i],vert$X75p_6x[i],vert$X75p_2y[i],vert$X75p_6y[i],vert$X75p_2z[i],vert$X75p_6z[i])
  
  #intervertebral sapce in bl
  dist.output$ivs_25_std[i] <- std.by.length(dist.output$ivs_25[i],dist.output$std_length[i])
  dist.output$ivs_50_std[i] <- std.by.length(dist.output$ivs_50[i],dist.output$std_length[i])
  dist.output$ivs_75_std[i] <- std.by.length(dist.output$ivs_75[i],dist.output$std_length[i])
  
}

```

# Questions
I took the table output in the above chunk and reorganized so that each measurment has it's own column with a new column noting the position along the vertebral column (25,50, or 75). To import this dataset run the Import Data chunk.

### Does centrum length differ by species (no: F=1.057, p=0.39), along spinal column (yes:F=5.256, p=0.009), or among climbing styles (no: F=1.813, p=0.176)?
```{r centrum length}
#Standardized centrum lenght does not differ by species F=1.057, p=0.39
cl_species = aov(cl_std ~ Species, data = GobyData)
summary(cl_species)

#Std CL does not differ by climbing style F=1.813, p=0.176
cl_style = aov(cl_std ~ style, data = GobyData)
summary(cl_style)

#Std CL does differ by postion along the vertebral column F=5.256, p=0.009
cl_pos = aov(cl_std ~ position, data = GobyData)
summary(cl_pos)

TukeyHSD(cl_pos) #posthoc for postion along vertebral column

#Plot
gg <- ggplot(data=GobyData, aes(x=position, y=cl_std)) +
  geom_violin() +
  geom_boxplot(width=0.2) +
  ylab("Centrum length (body lengths)") +
  xlab("Position along vertebral column (%)") +
  theme_classic() +
  theme(text=element_text(size=14),plot.title=element_text(hjust=0), legend.position = "none") 
gg
```

### Does minimum centrum width differ by species (yes:F=3.597, p=0.0135), along spinal column (yes:F=6.115, p=0.00467), or among climbing styles (yes: F=3.6, p=0.0361)?
```{r minimum centrum width}
#Minimum centrum width does differ by species F=3.597, p=0.0135
minCW_species = aov(can.d_std ~ Species, data = GobyData)
summary(minCW_species)

TukeyHSD(minCW_species) #posthoc

#minCW does differ by climbing style F=3.6, p=0.0361
minCW_style = aov(can.d_std ~ style, data = GobyData)
summary(minCW_style)

TukeyHSD(minCW_style) #posthoc

#minCW does differ by postion along the vertebral column F=6.115, p=0.00467
minCW_pos = aov(can.d_std ~ position, data = GobyData)
summary(minCW_pos)

TukeyHSD(minCW_pos) #posthoc

#colors
grey <- rgb(127, 127, 127, max = 255)
blue <- rgb(91, 155, 213, max = 255)
orange <- rgb(237, 125, 49, max = 255)
c_sch <- c(grey, blue, orange)

GobyData$position <- factor(GobyData$position, levels = c("25", "50", "75"), 
                  labels = c("25%*", "50%", "75%"))

#Plot
gg <- ggplot(data=GobyData, aes(x=Species, y=can.d_std)) +
  geom_violin(aes(fill=style)) + 
  geom_boxplot(width = 0.1) +
  #geom_point(position=pd,size=5) +
  #geom_errorbar(aes(ymin= cent.min.w$average - cent.min.w$std.error, ymax = cent.min.w$average + cent.min.w$std.error), width=0.25, position = pd)+
  ylab("Minimum centrum width (body lengths)") +
  scale_x_discrete("Species", limits = c("Eleotris sandwicensis", "Stenogobius hawaiiensis", "Awaous stamineus",  "Lentipes concolor", "Sicyopterus stimpsoni"), 
                   labels=expression(paste(italic("E.s.")), paste(italic("S.h.")), paste(italic("A.s.")), paste(italic(bold("L.c.*"))), paste(italic(bold("S.s.*")))))+
  scale_fill_manual(values = c_sch, name= 'Climbing Style', breaks= c("Non-climbing", "Powerburst", "Inching"), labels=expression(paste("Non-climbing"), paste(bold("Powerburst*")), paste(bold("Inching*"))))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
    panel.background = element_blank(), axis.line = element_line(colour = "black"), legend.key = element_rect(fill = NA),    
    plot.margin=unit(x=c(0.1,0.1,0.1,0.1),'cm'),text=element_text(size=14),plot.title=element_text(hjust=0), legend.position = "none") +
  facet_wrap(.~position) + 
  theme(strip.text.x = element_text(size=18, face="bold"))
gg
```


### Does maximum centrum width differ by species (no:F=2.45, p=0.062), along spinal column (no:F=60.62, p=0.55), or among climbing styles (yes:F=6.115, p=0.00467)?
```{r maximum centrum width}
#Maximum centrum width does not differ by species F=2.45, p=0.062
maxCW_species = aov(con.d_std ~ Species, data = GobyData)
summary(maxCW_species)

#maxCW does differ by climbing style F=3.27, p=0.048
maxCW_style = aov(con.d_std ~ style, data = GobyData)
summary(maxCW_style)

TukeyHSD(maxCW_style) #posthoc

#maxCW does not differ by postion along the vertebral column F=0.62, p=0.55
maxCW_pos = aov(con.d_std ~ position, data = GobyData)
summary(maxCW_pos)

grey <- rgb(127, 127, 127, max = 255)
blue <- rgb(91, 155, 213, max = 255)
orange <- rgb(237, 125, 49, max = 255)
c_sch <- c(grey, blue, orange)

#Plot
gg <- ggplot(data=GobyData, aes(x=style, y=can.d_std)) +
  geom_violin(aes(fill=style)) +
  geom_boxplot(width=0.2) +
  ylab("Maximum centrum width (body lengths)") +
  scale_x_discrete("Climbing Style", limits= c("Non-climbing", "Powerburst", "Inching"), labels=expression(paste("Non-climbing"), paste(bold("Powerburst*")), paste(bold("Inching*")))) +
  scale_fill_manual(values = c_sch)+
  theme_classic() +
  theme(text=element_text(size=14),plot.title=element_text(hjust=0), legend.position = "none") 
gg
```

### Does intervertebral space differ by species (yes: F=4.92, p=0.0025), along spinal column (no:F=1.1, p=0.34), or among climbing styles (yes:F=7.7, p=0.0014)?
```{r intervertebral space}
#Intervertebral space does differ by species F=4.92, p=0.0025
ivs_species = aov(ivs_std ~ Species, data = GobyData)
summary(ivs_species)

TukeyHSD(ivs_species) #posthoc

#ivs does differ by climbing style F=7.7, p=0.0014
ivs_style = aov(ivs_std ~ style, data = GobyData)
summary(ivs_style)

TukeyHSD(ivs_style) #posthoc

#ivs does not differ by postion along the vertebral column F=1.1, p=0.34
ivs_pos = aov(ivs_std ~ position, data = GobyData)
summary(ivs_pos)

#import pooled data table for ivs
pooled.ivs <- data.table(read.table('pooled indiv ivs.txt',header=TRUE,sep='\t'))
pooled.ivs$per.length <- factor(pooled.ivs$per.length, levels = c("25", "50", "75"), 
                  labels = c("25%", "50%", "75%"))

#colors
grey <- rgb(127, 127, 127, max = 255)
blue <- rgb(91, 155, 213, max = 255)
orange <- rgb(237, 125, 49, max = 255)
c_sch <- c(grey, blue, orange)

#Plot
gg <- ggplot(data=GobyData, aes(x=Species, y=ivs_std)) +
  geom_violin(aes(fill=style)) + 
  geom_boxplot(width = 0.2) +
  ylab("Intervertebral joint (body lengths)") +
  scale_x_discrete("Species", limits = c("Eleotris sandwicensis", "Stenogobius hawaiiensis", "Awaous stamineus",  "Lentipes concolor", "Sicyopterus stimpsoni"), 
                   labels=expression(paste(italic(bold("E.s.*"))), paste(italic(bold("S.h.*"))), paste(italic("A.s.")), paste(italic(bold("L.c.*"))), paste(italic(bold("S.s.**")))))+
  scale_fill_manual(values = c_sch, name= 'Climbing Style', breaks= c("Non-climbing", "Powerburst", "Inching"), labels=expression(paste(bold("Non-climbing*")), paste("Powerburst"), paste(bold("Inching*"))))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
    panel.background = element_blank(), axis.line = element_line(colour = "black"), legend.key = element_rect(fill = NA),    
    plot.margin=unit(x=c(0.1,0.1,0.1,0.1),'cm'),text=element_text(size=14),plot.title=element_text(hjust=0), legend.position = "none")
gg
```

